---
title: "Preprocessing of Witcher et al. TBI microglia"
author: "James Choi"
date: "Last compiled:`r Sys.Date()`"
output: 
  html_document
editor_options:
  chunk_output_type: 
    console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE, fig.align='center', 
                      tidy.opts=list(width.cutoff=80), tidy=TRUE,
                      results='hold')
```

```{r}
if(!grepl('scripts', getwd())) {setwd('scripts')}
results_out <- '../results/GSE160763_Witcher/'
data_out <- '../data/GSE160763_Witcher/'
dir.create(path = results_out)
dir.create(path = data_out)
```


Data was provided by the following:

Title: Traumatic Brain Injury Causes Chronic Cortical Inflammation and Neuronal Dysfunction Mediated by Microglia

Citation: Kristina G. Witcher, Chelsea E. Bray, Titikorn Chunchai, Fangli Zhao, Shane M. O'Neil, Alan J. Gordillo, Warren A. Campbell, Daniel B. McKim, Xiaoyu Liu, Julia E. Dziabis, Ning Quan, Daniel S. Eiferman, Andy J. Fischer, Olga N. Kokiko-Cochran, Candice Askwith, Jonathan P. Godbout. Traumatic Brain Injury Causes Chronic Cortical Inflammation and Neuronal Dysfunction Mediated by Microglia. Journal of Neuroscience 17 February 2021, 41 (7) 1597-1616; DOI:10.1523/JNEUROSCI.2469-20.2020

[Link to paper](https://doi.org/10.1523/JNEUROSCI.2469-20.2020)


```{r download_decompress}
dlcheck <- download.file(
  url = 'https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE160763&format=file',
  destfile = paste0(data_out, 'GSE160763_RAW.tar'),
  mode = 'wb' # the TAR file is a binary
)
if (dlcheck != 0) {
  stop('GSE160763 download unsuccessful')
} else {
  untar(tarfile = paste0(data_out, 'GSE160763_RAW.tar'), 
        exdir = paste0(data_out, 'GSE160763_RAW/'))
}
wit_files <- list.files(
  path = paste0(data_out, 'GSE160763_RAW/'), 
  full.names = TRUE
)
counts_out <- paste0(data_out, 'GSE160763_RAW/') # update data out folder
```

Set directory structure appropriate for `Seurat::Read10X()`.

```{r set_data_directories}
tmp <- sapply(strsplit(wit_files, split = '/'), `[`, 5)
sample_names <- unique(sapply(
  X = strsplit(tmp, '_'),
  FUN = function(x) paste(x[1:3], collapse = '_')
))
for (i in 1:length(sample_names)) {
  tmp_dir <- paste0(counts_out, sample_names[i], '/')
  dir.create(path = tmp_dir)
  tmp_files <- wit_files[grepl(sample_names[i], wit_files)]
  for (j in 1:length(tmp_files)) {
    tmp_from <- tmp_files[j]
    tmp_to <- unlist(strsplit(tmp_files[j], '/'))
    tmp_to <- tmp_to[length(tmp_to)]
    tmp_to <- unlist(strsplit(tmp_to, split = '_'))
    tmp_to <- tmp_to[length(tmp_to)]
    tmp_to <- paste0(tmp_dir, tmp_to)
    file.rename(from = tmp_from, to = tmp_to)
  }
}
rm(list = ls()[grepl('tmp', ls())])
```


From GEO:

>Overall design: Mice were subjected to midline fluid percussion injury (TBI) or kept as naïve controls. Half of the mice were fed PLX5622 (colony stimulating factor 1-receptor antagonist) in the chow prior to injury and up until the 7 dpi endpoint; the other half were fed AIN-7 vehicle chow. PLX5622 and CSF1R antagonism were used to deplete microglia. Therefore there were four experimental groups: Control, Control+PLX5622, TBI, and TBI+PLX5622. Cortex was dissociated and single cell suspensions from 3 mice per group were pooled into a single sample. This experiment was completed in two replicates (6 mice per group total; 2 experimental replicates per group).


Group-sample name key:

* N: normal (control)
* T: traumatic brain injury (TBI)
* P: PLX5622 treated (Plexxikon; microglia depletion)


```{r import_wt_data}
# Remove plexxikon animals
sample_names <- list.dirs(counts_out, recursive = FALSE)
sample_names <- sample_names[!grepl('P_', sample_names)]
counts <- vector(mode = 'list', length = length(sample_names))

for (i in 1:length(sample_names)) {
  counts[[i]] <- Seurat::Read10X(
    data.dir = sample_names[i],
    strip.suffix = TRUE, 
    gene.column = 2
  )
}
```

Set unique cell barcodes to track back to original study sample.

```{r unique_barcodes}
barcode_append <- strsplit(sample_names, '/')
barcode_append <- sapply(barcode_append, FUN = function(x) x[length(x)])
barcode_append <- gsub(pattern = '_', replacement = '.', x = barcode_append)
if (length(barcode_append) != length(counts)) {
  stop('Number of barcode appends not equal to number of count matrices')
}
for (i in 1:length(counts)) {
  colnames(counts[[i]]) <- paste(colnames(counts[[i]]),
                                 barcode_append[i],
                                 sep = '_')
}
```

```{r seurat_objects}
library('Seurat')
genes <- lapply(X = counts, FUN = rownames)
if (!all(sapply(genes[2:length(genes)], identical, genes[[1]]))) {
  stop('Not all count matrices have same genes')
} else {
  counts <- Reduce(f = cbind, x = counts)
}
# dim(counts)

# Set up metadata and experimental details
metadata <- data.frame(
  'study' = 'Witcher2021',
  'sample_id' = sapply(strsplit(colnames(counts), '_'), `[`, 2),
  'tissue' = 'cortex',
  'injury_model' = 'TBI'
)
rownames(metadata) <- colnames(counts)
tmp_inj <- sapply(strsplit(x = metadata$sample_id, split = '\\.'), `[`, 2)
tmp_inj <- plyr::mapvalues(tmp_inj, c('N','T'), c('mFPI.Ctrl', 'mFPI.7dpi'))
metadata$injury <- tmp_inj

Witcher2021 <- list(counts = counts, metadata = metadata)
saveRDS(Witcher2021, file = paste0(data_out, 'Witcher2021.rds'))
```

