---
title: "Preprocessing of Lee et al. Tauopathy/AD Microglia"
author: "James Choi"
date: "Last compiled:`r Sys.Date()`"
output: 
  html_document
editor_options:
  chunk_output_type: 
    console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE, fig.align='center', 
                      tidy.opts=list(width.cutoff=80), tidy=TRUE,
                      results='hold')
```

```{r}
if(!grepl('scripts', getwd())) {setwd('scripts')}
results_out <- '../results/GSE153895_Lee/'
data_out <- '../data/GSE153895_Lee/'
dir.create(path = data_out)
```


GEO Accession: GSE153895

Data was provided by the following:

Title: Trem2 restrains the enhancement of tau accumulation and neurodegeneration by β-amyloid pathology.

Citation: Seung-Hye Lee, William J. Meilandt, Luke Xie, Vineela D. Gandham, Hai Ngu, Kai H. Barck, Mitchell G. Rezzonico, Jose Imperio, Guita Lalehzadeh, Melanie A. Huntley, Kimberly L. Stark, Oded Foreman, Richard A.D. Carano, Brad A. Friedman, Morgan Sheng, Amy Easton, Christopher J. Bohlen, David V. Hansen.  Trem2 restrains the enhancement of tau accumulation and neurodegeneration by β-amyloid pathology. Neuron, Volume 109, Issue 8, 2021, Pages 1283-1301.e6, 
ISSN 0896-6273, https://doi.org/10.1016/j.neuron.2021.02.010.

[Link to paper](https://doi.org/10.1016/j.neuron.2021.02.010)


```{r download_decompress}
dlcheck <- download.file(
  url = 'https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE153895&format=file',
  destfile = paste0(data_out, 'GSE153895_RAW.tar'),
  mode = 'wb' # the TAR file is a binary
)
if (dlcheck != 0) {
  stop('GSE153895 download unsuccessful')
} else {
  untar(tarfile = paste0(data_out, 'GSE153895_RAW.tar'), 
        exdir = paste0(data_out, 'GSE153895_RAW/'))
}
dlcheck <- download.file(
  url = 'https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE153895&format=file&file=GSE153895%5FTsneCoordinatesAndClusterAssignments%2Etxt%2Egz',
  destfile = paste0(data_out, 'GSE153895_TsneCoordinatesAndClusterAssignments.txt.gz'),
  mode = 'wb' # the TAR file is a binary
)
if (dlcheck != 0) {
  stop('GSE153895_TsneCoordinatesAndClusterAssignments.txt.gz download unsuccessful')
} else {
  sample_metadata <- read.table(
    file = gzfile(paste0(data_out, 'GSE153895_TsneCoordinatesAndClusterAssignments.txt.gz')),
    sep = '\t',
    header = TRUE
  )
}
```


From GEO:

> Overall design: RNA isolated from sorted hippocampi from male non-transgenic (n=2), Trem2ko (n=1), P301Lhomo (n=3), TauPS2APP (n=3), and TauPS2APP;Trem2ko (n=3) mice. All mice were 19-22 months at the time of perfusion. Following the manuscript, the one Trem2ko mouse was analyzed as a control, as indicated in the "genotype group" sample characteristic.

I will use the P301L, PS2/APP/P301L, and Non-transgenic mouse data.

```{r import_wt_data}
# Select genotypes
tmp_keep <- c('NonTg', 'P301L','PS2/APP/P301L')
tmp_map <- apply(
  X = table(sample_metadata$sampleID, sample_metadata$genotype),
  MARGIN = 1,
  FUN = function(x) names(which(x > 0))
)
sample_map <- tmp_map[tmp_map %in% tmp_keep]
tmp_keep <- paste(names(sample_map), collapse = '|')

# extract counts for matching files
sample_files <- list.files(
  path = paste0(data_out, 'GSE153895_RAW/'), 
  full.names = TRUE
)
sample_files <- sample_files[grepl(pattern = tmp_keep, x = sample_files)]
counts <- vector(mode = 'list', length = length(sample_files))
for (i in 1:length(sample_files)) {
  message('Importing:', sample_files[i])
  tmp_counts <- read.table(
    file = gzfile(sample_files[i]),
    sep = '\t',
    header = TRUE,
    row.names = 1
  )
  tmp_counts <- as.matrix(tmp_counts)
  counts[[i]] <- Matrix::Matrix(data = tmp_counts, sparse = TRUE)
}
rm(list = ls()[grep('tmp', ls())])
```

Set unique cell barcodes to track back to original study sample.

```{r unique_barcodes}
sample_names <- strsplit(sample_files, '/')
sample_names <- sapply(sample_names, function(x) x[length(x)])
sample_names <- sapply(strsplit(sample_names, '_'), `[`, 1)
for (i in 1:length(counts)) {
  tmp_barcodes <- plyr::mapvalues(
    x = colnames(counts[[i]]),
    from = sample_metadata$cellID,
    to = sample_metadata$barcode,
    warn_missing = FALSE
  )
  colnames(counts[[i]]) <- paste(tmp_barcodes, sample_names[i], sep = '_')
}
```

Setup metadata including experimental details.

```{r merge_counts}
genes <- lapply(X = counts, FUN = rownames)
if (!all(sapply(genes[2:length(genes)], identical, genes[[1]]))) {
  shared_genes <- Reduce(f = intersect, x = genes)
  counts <- lapply(X = counts, FUN = function(x) x[shared_genes,])
  counts <- Reduce(f = cbind, x = counts)
} else {
  counts <- Reduce(f = cbind, x = counts)
}
dim(counts)
```

Setup metadata including experimental details.

```{r metadata}
# Set up metadata and experimental details
metadata <- data.frame(
  'study' = 'Lee2021',
  'sample_id' = sapply(strsplit(colnames(counts), '_'), `[`, 2),
  'tissue' = 'hippocampus',
  'injury_model' = 'mouseAD'
)
rownames(metadata) <- colnames(counts)

tmp_name <- strsplit(sample_files, '/')
tmp_name <- sapply(tmp_name, function(x) x[length(x)])
tmp_name <- sapply(strsplit(tmp_name, split = '\\.'), `[`, 1)
tmp_map <- sapply(strsplit(tmp_name, '_'), `[`, 1)
names(tmp_map) <- sapply(strsplit(tmp_name, '_'), `[`, 2)
sample_metadata$GSE <- plyr::mapvalues(
  x = sample_metadata$sampleID,
  from = names(tmp_map),
  to = tmp_map
)
sample_metadata$long_barcode <- paste(
  sample_metadata$barcode,
  sample_metadata$GSE,
  sep = '_'
)
tmp_inj <- plyr::mapvalues(
  x = colnames(counts),
  from = sample_metadata$long_barcode,
  to = sample_metadata$genotype,
  warn_missing = FALSE
)
metadata$injury <- tmp_inj
metadata$injury <- gsub('/', '-', metadata$injury)
metadata$injury <- plyr::mapvalues(
  x = metadata$injury,
  from = c('NonTg','P301L','PS2-APP-P301L'),
  to = c('Aged.Ctrl', 'Aged.P301L', 'Aged.PS2-APP-P301L')
)
metadata$age <- 'P600'
metadata$sex <- 'M'
```

Save

```{r save}
Lee2021 <- list(counts = counts, metadata = metadata)
saveRDS(Lee2021, file = paste0(data_out, 'Lee2021.rds'))
```

```{r}
sessionInfo()
```

